---
  title: "Geospatial Project"
  description: "EDS223 Final Project"
  author:
    - name: Maxwell Patterson
    - affiliation: Master of Environmental Data Science Program
    - affiliation-url: https://ucsb-meds.github.io/
  date: 12-01-2023
  categories: [R, MEDS, Maps]
  bibliography: references.bib
---

# A Geospatial Analysis of Houston's Blackout Sensitivity

## Introduction

In the heart of winter, the city of Houston found itself plunged into darkness. The February 2021 blackout, a consequence of severe winter storms, left a metropolis in a precarious state, exposing the vulnerabilities in urban power grids. This blog post aims to dissect the Houston blackout, employing geospatial analysis to estimate the number of homes affected and explore the socio-economic factors influencing community resilience during power outages. This intricate analysis is not just a tale of a city in darkness but a lens into the fragility of our urban lifelines.

## Background

Urban blackouts are more than mere inconveniences; they're a stark reminder of our dependence on stable power sources. In Houston, a city with a complex infrastructure and a propensity for extreme weather events, the susceptibility to blackouts raises critical concerns about preparedness and resilience. A study by the U.S. Energy Information Administration highlights Houston's energy landscape, noting its significant role in national energy supply but also its vulnerability to disruptions. This backdrop sets the stage for our analysis of the 2021 blackout, a pivotal moment that not only challenged the city's infrastructure but also tested the resilience of its communities.

## Data and Methodology

The analysis hinges on three key data sources: VIIRS night light data, OpenStreetMap data detailing roads and buildings, and American Community Survey (ACS) socio-economic data. We employed R, utilizing libraries such as dplyr, sf, and terra, to manipulate and analyze these datasets. Our methodology involved a step-by-step geospatial approach, beginning with the aggregation of night light intensity data to assess blackout areas. We then integrated this with road and building data to estimate the number of homes affected and concluded with an examination of socio-economic factors, offering insights into the recovery dynamics of different communities.

Let's get to coding.

Starting off by loading in the necessary libraries.

```{r, message=FALSE}
# import libraries
library(dplyr)
library(abind)
library(sf)
library(terra)
library(spData)
library(spDataLarge)
library(tidyverse)
library(dplyr)
library(stars)
library(here)

# clear environment for sanity
rm(list = ls())
```

Next, importing data. The first set of data is the night light data, tiles or images that capture the light intensity emanating from the surface of the planet during nighttime. The data is stored as rasters, and the `stars` package is used for raster handling.

```{r}
# read in night lights tiles
feb7_h08v05 <- read_stars(here::here("blog", "2023-12-2-geo-proj", "data", "VNP46A1", "VNP46A1.A2021038.h08v05.001.2021039064328.tif"))
feb7_h0806 <- read_stars(here::here("blog", "2023-12-2-geo-proj", "data", "VNP46A1", "VNP46A1.A2021038.h08v06.001.2021039064329.tif"))
feb16_h0805 <- read_stars(here::here("blog", "2023-12-2-geo-proj", "data", "VNP46A1", "VNP46A1.A2021047.h08v05.001.2021048091106.tif"))
feb16_h0806 <- read_stars(here::here("blog", "2023-12-2-geo-proj", "data", "VNP46A1", "VNP46A1.A2021047.h08v06.001.2021048091105.tif"))
```

Next, we want to combine the tiles that have the same date and create two objects. Let's do this and plot the tiles for each date for inspection purposes. We can utilize the `st_mosaic()` function to combine the tiles.

```{r}
# combine tiles for each date to create 2 objects
feb7 <- st_mosaic(feb7_h08v05, feb7_h0806)
feb16 <- st_mosaic(feb16_h0805, feb16_h0806)
```

Great. The next step involves creating a blackout mask to identify areas in Houston that lost power during the winter storms. This is done by calculating the change in light intensity, reclassifying this difference raster, then converting the raster to a stars object. ADD MORE.

Next, we want to convert the blackout stars to a vector format and fix any invalid geometries using `st_make_valid()`. Let's visualize the blackout in vector format to check where we are at.

```{r}
# calculate the change in lights intensity from Feb 7 to Feb 16
diff <- feb7 - feb16

# reclassify difference raster so it is maleable
diff <- rast(diff)
rmask <- diff

# reclassify difference raster assuming a 200 nW shift was due to the blackout
rmask[rmask <= 200] = NA

# convert back to a stars object 
blackout_stars <- st_as_stars(rmask)

# convert blackout to vector format and fix invalid geometries, set CRS
blackout_sf_valid_trans <- st_as_sf(blackout_stars, as_points = FALSE) %>% 
  st_make_valid() %>% 
  st_transform(crs = 3083)

# plot the blackout now that it is in vector format for inspection
plot(st_geometry(blackout_sf_valid_trans))
```

This visual shows the light intensity of the greater Houston region.

The next piece of analysis involves defining the Houston metropolitan area and cropping the blackout mask to this area of interest. It is imperative that the coordinate reference systems align of these two spatial objects!

```{r, warning=FALSE}
# define Houston metropolitan coordinates
houston_coords <- matrix(c(-96.5, 29, -96.5, 30.5, -94.5, 30.5, -94.5, 29, -96.5, 29), ncol = 2, byrow = TRUE)

# create Houston polygon outline
houston_poly <- st_polygon(list(houston_coords))

# convert to sf and initially assign CRS EPSG:4326
houston <- st_sfc(houston_poly, crs = st_crs("EPSG:4326"))

# transform the Houston polygon to EPSG:3083 
houston <- st_transform(houston, crs = "EPSG:3083")

# crop the blackout to the Houston metropolitan region
cropped_blackout_proj <- st_intersection(blackout_sf_valid_trans, houston) 
 
# assign EPSG:3083 CRS to assure consistent CRS
cropped_blackout_proj <- st_transform(cropped_blackout_proj, crs = "EPSG:3083")

plot(cropped_blackout_proj)
```

This is a visual of the Houston metropolitan area light intensity. ADD MORE!!!!!

Next, we will refine the blackout analysis by excluding areas near highways. This helps in ensuring that the blackout mask we created earlier accurately reflects residential or non-highway areas that lost power, rather than areas simply experiencing less traffic during the storms.

```{r, warning=FALSE}
# define SQL query
query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"

# load highway data
highways <- st_read("data/gis_osm_roads_free_1.gpkg", query = query, quiet = TRUE)

# reproject highway to EPSG:3083
highway_proj <- st_transform(highways, crs = 3083)

# identify areas within 200m of all highways, dissolve buffer, and convert to sf 
buffer <- highway_proj %>% 
  st_buffer(200) %>% 
  st_union() %>% 
  st_as_sf()

# exclude areas close to highways from blackout
blackout <- st_difference(cropped_blackout_proj, buffer)
```

#### Finding Impacted Homes

Next, we will look to find the number of homes that have been impacted by the blackouts. A buildings dataset will be loaded in and filtered for residential buildings. Then, we will filter for the homes that are in the blackout area and display this value.

```{r}
# define SQL query
query_buildings <- "
SELECT * 
FROM gis_osm_buildings_a_free_1
WHERE (type IS NULL AND name IS NULL)
OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')
"

# load buildings dataset using st_read and apply the query, reproject CRS
buildings_proj <- st_read("data/gis_osm_buildings_a_free_1.gpkg", query = query_buildings, quiet = TRUE) %>% 
  st_transform(crs = 'EPSG:3083')

# filter for homes within blackout areas by filtering
impacted_homes <- buildings_proj[blackout, , op = st_intersects]

# count the number of impacted homes
num_impacted_homes <- nrow(impacted_homes)

# show the number of impacted homes
print(paste("Number of impacted homes:", num_impacted_homes))
```

## Results and Visualizations

-   **Blackout Impact Analysis**:
    -   Show maps identifying areas affected by the blackout
    -   Discuss any interesting patterns or findings
-   **Socio-Economic Analysis**:
    -   Display the box plot of median incomes and discuss socio-economic disparities

## Conclusions 

-   Summarize the key findings of your analysis
-   Discuss the implications of these findings on urban planning and emergency management
-   Acknowledge any limitations of your study
-   Suggest potential areas for future research
